---
title: "Time Series"
author: "Justin Williams"
date: "5/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, warning=FALSE}
library(tidyverse)
library(fpp2)
library(ggfortify)
library(forecast)
library(data.table)
library(scales)
library(lubridate)
library(RColorBrewer)
library(viridis)
library(ggseas)
library(sf)
library(mapview)
```


## Overivew

This notebook will perform exploratory data analysis on both the air quality and COVID dataset.

### Load in data

Load in simple dataframes of air quality data and combine into one dataframe for 
time series analysis.

```{r load-air-simple-quality-data}
simple_aqs <- list.files(
  path = "/Users/justinwilliams/projects/air_quality_covid/data/simple_yearly/", 
  pattern = ".Rds", full.names = T) %>% 
  map_dfr(readRDS) %>% #combine into one df
  mutate(date_local = as.Date(date_local))

simple_aqs

saveRDS(object = simple_aqs,
        file = "./data/simple_yearly/simple_aqs.rds")
```

Load in all data frames separately for some EDA

```{r load-in-full-df}
# list of paths to dfs
filenames <- list.files(
  path = "/Users/justinwilliams/projects/air_quality_covid/data/",
  pattern = ".rds")

# create list to name dfs
df_names <- filenames %>% str_remove(".rds")

# loop through each name and to each df 
for (i in df_names) {
  filepath <- file.path("/Users/justinwilliams/projects/air_quality_covid/data/",
                        paste(i, ".rds", sep=""))
  assign(i, readRDS(filepath))
}

```

## EDA

### Bar plots

Plot NYC Open AQ data which is annual mean.

```{r nyc-aq-open-data-bar}
nyc_aq %>% 
  filter(start_date >= c("2017-01-01") &
           measure_info != "ppb" &
           startsWith(time_period, "Annual"))%>% 
  group_by(start_date, time_period, measure_info) %>% 
  summarise(mean_measurement = mean(data_value, na.rm = T)) %>% 
  arrange((start_date)) %>% 
  ggplot() +
    geom_bar(aes(y = mean_measurement, x = start_date),
             stat = "identity") +
  coord_flip()
```

 Mean per year based on EPA data.
 
```{r year-mean-aqs-data}
mean_aqs_year <-  simple_aqs %>% 
  group_by(year = year(date_local)) %>% 
  summarise(mean_per_year = mean(mean_measurement)) %>%
  ggplot() +
    geom_bar(aes(y = mean_per_year, x = year, fill = factor(year)),
           stat = "identity", alpha = 0.8) +
    coord_flip() +
    scale_fill_manual(values = brewer_pal(palette = 3)(7)[3:7]) +
    labs(y = "PM2.5 (ug/m3)", x = "", title = "Mean PM2.5 per Year", fill = "") +
    theme_minimal() +
    theme(plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5,
                                    size = 18),
          axis.text = element_text(size = 12),
          axis.title.x = element_text(size = 14),
          legend.position = "none")

ggsave(filename = "./images/mean_aqs_year.png",
       width = 11, height = 7,
       plot = mean_aqs_year)

mean_aqs_year
```

Median per year

```{r median-year}
simple_aqs %>% 
  group_by(year = year(date_local)) %>% 
  summarise(median_per_year = median(mean_measurement)) %>%
  ggplot() +
  geom_bar(aes(y = median_per_year, x = year),
           stat = "identity") +
  coord_flip()
```


### Grouped bar charts

Let's group bar charts by month and facet wrap with year to get some idea of seasonality.

```{r grouped-bar-year}
grouped_bar_year <-  simple_aqs %>% 
  mutate(year = year(date_local),
         month = month(date_local,
                       label = T,
                       abbr = T)) %>% 
  group_by(year, month) %>% 
  summarise(mean_month = mean(mean_measurement)) %>% 
  ggplot(aes(fill = factor(year))) +
    geom_bar(aes(x = month, y = mean_month),
             stat = "identity", alpha = 0.8) +
    facet_wrap(~year, scales = "free_y") +
    scale_fill_manual(values = brewer_pal(palette = 3)(7)[3:7]) +
    labs(x = "", y = "PM2.5 (ug/m3)", 
         title = "Mean PM2.5 per Month by Year",
         fill = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 8),
          axis.title.x = element_text(size = 14),
          legend.text = element_text(size = 14),
          legend.position = c(0.85, 0.2))

ggsave(filename = "./images/grouped_bar_year.png",
       width = 11, height = 7, 
       plot = grouped_bar_year)

grouped_bar_year
```

Let's group by month agnostic to year

```{r group-month-agnostic-year}
simple_aqs %>% 
  mutate(month = month(date_local,
                       label = T,
                       abbr = T)) %>% 
  group_by(month) %>% 
  summarise(mean_month = mean(mean_measurement)) %>% 
  ggplot() +
    geom_bar(aes(x = month, y = mean_month),fill = "#A63446",
             stat = "identity", alpha = 0.8) +
    labs(x = "", y = "PM2.5", 
         title = "Mean PM2.5 per Month (2017 - 2021)",
         fill = "")  +
    scale_fill_brewer(palette = "Set3") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5),
          legend.position = "none")
```
Median per month

```{r median-per-month}
median_per_month <-  simple_aqs %>% 
  mutate(month = month(date_local,
                       label = T,
                       abbr = T)) %>% 
  group_by(month) %>% 
  summarise(med_month = median(mean_measurement)) %>% 
  ggplot() +
    geom_bar(aes(x = month, y = med_month),fill = "#0C6291",
             stat = "identity", alpha = 0.8) +
    labs(x = "", y = "PM2.5 (ug/m3)", 
         title = "Median PM2.5 per Month (2017 - 2021)",
         fill = "")  +
    scale_fill_brewer(palette = "Set3") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 12),
          axis.title.y = element_text(size = 14),
          legend.position = "none")

ggsave(filename = "./images/median_month.png",
       width = 11, height = 7,
       plot = median_per_month)
```

### Mapping test sites

We have lat-long but need to create geometry from the coordinates.
Let's isolate one year of full dataset and grab the sites and names and create a smaller df we can use to 

```{r}
nyc_county_dailysummary_2017
```

```{r create-sf-object}
# get sites, county, site address and site number
nyc_sites_sf <- st_as_sf(nyc_county_dailysummary_2017,
         coords = c("longitude","latitude"),
         crs = 4326) %>% 
  st_transform(2263) %>% 
  group_by(county, site_address, site_number) %>% 
  summarise()

# get nyc geography
nyc_nta <- nyc_boundaries(
  geography = "nta"
) %>%
  st_transform(2263)

# join site with nta to get names in df
nyc_sites_sf_nta <- nyc_sites_sf %>% 
  st_join(nyc_nta)

# list of test sites with nta_name
options(knitr.kable.NA = '')
nyc_sites_sf_nta %>% 
  st_drop_geometry() %>% 
  select(NTA = nta_name, Borough = borough_name) %>% 
  knitr::kable(format = "simple",
    caption = "Neighborhood Tabulation Areas and County of Air Quality test sites")
```

Preview with mapview

```{r map-nyc-test-sites}
mapview_nyc_test_sites <- nyc_sites_sf_nta %>% 
  mapview(
    col.regions = "red",
    legend = T,
    layer.name = "NYC Air Quality Test Sites",
    alpha = 0.8,
    cex = 4.5
  ) +
  mapview(nyc_nta,
          col.regions = c("lightblue", "deepskyblue4",
                               muted("blue"),
                               muted("purple"), "darkblue"),
          zcol = "borough_name",
          layer.name = "Borough",
          legend.opacity = 0.8)

mapshot(mapview_nyc_test_sites,
        file = "./images/mapview_nta_test_sites.png")

mapview_nyc_test_sites
```

### Summary statistics

Let's generate some summary stats of the distributions by year.

```{r summary-stats}
simple_aqs %>% 
  mutate(year = year(date_local)) %>% 
  group_by(year) %>% 
  summarize(min = round(min(mean_measurement),2),
            q1 = round(quantile(mean_measurement, 0.25),2),
            median = round(median(mean_measurement),2),
            mean = round(mean(mean_measurement),2),
            q3 = round(quantile(mean_measurement, 0.75),2),
            max = round(max(mean_measurement),2),
            sd = round(sd(mean_measurement),2))
```

There is an extreme outlier in 2021. Let's try and isolate that particular record.

```{r 2021-outlier}
simple_aqs %>% 
  arrange(desc(mean_measurement)) %>% 
  head(5)
```

Looks like it was July 20th 2021. Google search revealed that this is the day smoke from the fires in the Western US and Canada made its way across the continent and caused high air pollution on the each coast.
https://www.nytimes.com/2021/07/20/us/wildfire-smoke-new-york-city.html

Let's look at dates around that time.

```{r wildfires-air-pollution-july-2021}
simple_aqs %>% 
  filter(date_local >= "2021-07-19" &
           date_local <= "2021-07-31")
```

Looks like it was pretty much a one day event things kind of went back to normal thereafter. Maybe it came back around July 27th? But other then that looks normal.

### Histograms

Let's look at some of the distributions of PM2.5 by year

```{r hist-pm2.5-year}
hist_year <- simple_aqs %>% 
  mutate(year = year(date_local)) %>%
  group_by(year) %>% 
  ggplot(aes(x = mean_measurement,fill = factor(year,))) +
    geom_histogram(position = "identity", binwidth = 0.7,
                   alpha = 0.8) +
    facet_wrap(~year, scales = "free") +
    scale_fill_manual(values = brewer_pal(palette = 3)(7)[3:7]) +
    theme_minimal() +
    labs(x = "PM2.5 (ug/m3)", y = "Count", 
         title = "Distribution of Mean PM2.5 Year",
         fill = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 8),
          axis.title.x = element_text(size = 14),
          legend.text = element_text(size = 14)) +
    theme(legend.position = c(0.85, 0.2))

ggsave(filename = "./images/hist_mean_year.png",
       width = 11, height = 7,
       plot = hist_year)

hist_year
```
This shows the outrageous outlier in 2021, is this an error? Let;s confirm with some boxplots. Otherwise pretty normal.

Will prob need to zoom in on each individually to see how normal they really are, but they all seem skewed to the right. 

### Boxplots

Side by side boxplots separated by year. 

```{r boxplots}
simple_aqs %>% 
  mutate(year = year(date_local)) %>%
  group_by(year) %>% 
  ggplot(aes(x = mean_measurement,fill = factor(year,))) +
    geom_boxplot(position = "identity",
                   alpha = 0.8) +
    facet_wrap(~year, scales = "free") +
    scale_fill_manual(values = brewer_pal(palette = 3)(7)[3:7]) +
    theme_minimal() +
    labs(x = "", y = "", fill = "") +
    theme_minimal() +
    theme(legend.position = c(0.85, 0.2)) +
    coord_flip()
```

Look at this without `facet_wrap()`.

```{r boxplot-no-facet}
box_year <- simple_aqs %>% 
  mutate(year = year(date_local)) %>%
  group_by(year) %>% 
  ggplot(aes(y = mean_measurement, x = year,fill = factor(year,))) +
    geom_boxplot(position = "identity",
                   alpha = 0.8) +
    scale_fill_manual(values = brewer_pal(palette = 3)(7)[3:7]) +
    theme_minimal() +
    labs(x = "", y = "PM2.5 (ug/m3)", 
         title = "Mean PM2.5 per year",fill = "") +
    theme_minimal() +
    theme(legend.position = c(0.1, 0.85),
          plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 14),
          axis.title.x = element_text(size = 16),
          legend.text = element_text(size = 14))

ggsave(filename = "./images/box_year.png",
       width = 11, height = 7,
       plot = box_year)

box_year
```

Let's isolate 2021

```{r 2021-box}
simple_aqs %>% 
  mutate(year = year(date_local)) %>%
  group_by(year) %>%
  filter(year == 2021) %>% 
  ggplot(aes(x = mean_measurement,fill = factor(year,))) +
    geom_boxplot(position = "identity", binwidth = 0.7,
                   alpha = 0.8) +
    theme_minimal() +
    coord_flip() +
    theme(legend.position = "None")

```


### Line plots

Let's do a simple line plot

```{r simple-plot}
# 3 years of data 2017:2019
plot(simple_aqs[1:1095,], type = "l")
```
Looks like there is a slight trend, some seasonality and a bunch of randomness.
Let's make this a time series and re-plot.

```{r create-ts-object}
ts_nyc_2 <- ts(simple_aqs$mean_measurement, start = decimal_date(as.Date("2017-01-01")),
               frequency = 365)

# save time series object
saveRDS(ts_nyc_2,
        file = "./data/ts_nyc.rds")
```

Let's replot `ts()` object

```{r ts-plot}
plot(ts_nyc_2, start = c(2017,01), end = c(2019, 12))
ts_nyc_2
```

Try this in ggplot

```{r ggplot-line}
line_all <- simple_aqs %>% 
  ggplot(aes(x = date_local, y = mean_measurement)) + 
    geom_line(color = "#0C6291", alpha = 0.8) +
    theme_minimal() +
    labs(x = "", y = "PM2.5 (ug/m3)", 
         title = "Mean PM2.5 per Day",
         color = "") +
    theme_minimal() +
    theme(plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 14),
          axis.title.x = element_text(size = 16))

ggsave(filename = "./images/line_all.png",
       width = 11, height = 7, plot = line_all)

line_all
```


Separate by year.

```{r by-year}
ggseasonplot(ts_nyc_2,
             year.labels = T,
             continuous = T,
             xlab = "Time",
             ylab = "PM2.5",
             main = "PM2.5 by year") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme_minimal()
```
GGseasonal plot doesn't seem to let me customize x-axis date breaks. 
Let's try plotting something like this with the dateframe.

```{r ggplot-line-chart-group-year}
simple_aqs %>%
  mutate(year = year(date_local),
         month = month(date_local,
                       label = T,
                       abbr = T)) %>% 
  group_by(year, month) %>% 
  summarise(mean_month = mean(mean_measurement)) %>% 
  ggplot(aes(month, mean_month, 
             group = factor(year),
             color = factor(year))) +
    geom_line(lwd = 2) +
    scale_color_manual(values = brewer.pal(6,"Dark2")) +
    theme_minimal() +
    labs(x = "", y = "PM2.5 (ug/m3)", 
         title = "Mean PM2.5 per Month by Year",
         color = "") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = c(0.91, 0.8))

```


Let's zoom in on July 2021 to get a visual of the spike.

```{r line-plot-july-2021}
july_2021 <- simple_aqs %>% 
  ggplot(aes(x = date_local, y = mean_measurement)) +
    geom_line(color = "#0C6291",
              lwd = 1, alpha = 0.8) +
    geom_point(size = 2) +
    theme_minimal() +
    labs(x = "", y = "PM2.5 (ug/m3)", title = "July 2021 Mean PM2.5 per Day") +
    theme(axis.text.x = element_text(angle = 60, hjust = 1),
          plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 14),
          axis.title.x = element_text(size = 16)) +
    scale_x_date(limit = c(as.Date("2021-07-01"), as.Date("2021-07-31"))) 

ggsave(filename = "./images/july_2021.png",
       width = 11, height = 7,
       plot = july_2021)

july_2021
```


### Decompose

Let's try out decompose

```{r decompose-ts-object}
nyc_decompose_ts_2 <- decompose(ts_nyc_2)
plot(nyc_decompose_ts_2)
```

Let's use the **ggseas** package to get a better looking version.

```{r ggseas-decompose}
decompose_ggplot <-  simple_aqs %>%
  ggsdc(aes(date_local,
            mean_measurement),
        frequency = 365,
        
        method = "decompose",
        facet.titles = c("Original Series",
                         "Underying Trend",
                         "Seasonal patterns",
                         "Residual Randomness")) +
    geom_line() +
    theme_minimal() +
    labs(x = "Year", y="", title = "Decomposition of PM2.5 (ug/m3) 2017 - 2021") +
    theme(plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          strip.text = element_text(size = 14),
          axis.text.x = element_text(size = 12))

ggsave(filename = "./images/decompose_ggplot.png",
       width = 11, height = 7,
       plot = decompose_ggplot)

decompose_ggplot
```


Trend varies, but looks like it starts trending down once 2020 hits, then goes back up in 2021 with a huge spike towards the summer (extreme wildfires). There does seem to be significant seasonality, spiking in the summer and winter. 

Let's try and adjust for seasonality.

```{r adjust-seasonality-by-year}

# aggregate by month
month_avg <- simple_aqs %>% 
  mutate(year = year(date_local),
         month = month(date_local,
                       label = T,
                       abbr = T)) %>% 
  group_by(year, month) %>% 
  summarise(mean_month = mean(mean_measurement))

# create ts object
month_avg_ts <-  ts(month_avg, frequency = 12, start = c(2017,01))

# convert back to df to get floating datetime
month_avg_df <- tsdf(month_avg_ts)

# plot with seasonality adjustment
seas_adj_month <- month_avg_df %>% 
  ggplot(aes(x, mean_month)) + 
    geom_point() +
    stat_seas(col = "#0C6291",
              lwd = 1, alpha = 0.8) +
    theme_minimal() +
    labs(x = "", y="PM2.5 (ug/m3)", title = "Seasonally Adj Monthly Mean PM2.5") +
    theme(plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 12),
          axis.title.y = element_text(size = 12))

ggsave(filename = "./images/seas_adj_month.png",
       width = 11, height = 7,
       plot = seas_adj_month)

seas_adj_month
```

### Rolling Average

General rolling average graph with ggplot.

```{r rolling-averages}
rolling_avg_ggplot <- simple_aqs %>% 
  ggplot(aes(x = date_local, y = mean_measurement)) +
    geom_line(color = "grey75") +
    stat_rollapplyr(width = 30, 
                    align = "center",
                    lwd = 1) +
    labs(x = "", y = "PM2.5 (ug/m3)", title = "Rolling Average PM2.5 by Year",
         color = "") +
    theme_minimal() +
    theme(plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 10),
          axis.title.y = element_text(size = 10)) +
    theme(legend.position = c(0.85, 0.2))

ggsave(filename = "./images/rolling_avg_ggplot.png",
       width = 11, height = 7,
       plot = rolling_avg_ggplot)

rolling_avg_ggplot
```


### NYC Covid data EDA

Some brief EDA on NYC Covid data to get an idea of rise and fall of COVID cases in 2020 and 2021.

#### Line Plot

```{r covid-line-plot}
covid_cases <- nyc_simple %>% 
  ggplot(aes(x = date, y = as.numeric(cases))) + 
    geom_line(color = "#0C6291", alpha = 0.8) +
    scale_y_continuous(labels = scales::number_format(scale = .0001,
                                                     suffix = "k")) +
    theme_minimal() +
    labs(x = "", y = "Case Count", 
         title = "COVID Case Count per Day",
         color = "") +
    theme_minimal() +
    theme(plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 14),
          axis.title.y = element_text(size = 16))

ggsave(filename = "./images/nyc_covid_casaes.png",
       width = 11, height = 7, plot = covid_cases)

covid_cases
```

Let's segment the aqs data then transform both series to view on same line plot.

### Compare time series

```{r segment-log-transform-data}
# segment data
seg_aqs <- simple_aqs %>% 
  filter(date_local >= as.Date("2020-02-29")) %>% 
  rename(date = date_local)

seg_covid <- nyc_simple %>% 
  mutate(cases = as.numeric(cases),
         date = as.Date(date)) %>% 
  filter(date <= as.Date("2021-12-31")) %>% 
  select(date, cases)

# take log transformation
seg_aqs_log <- log(seg_aqs$mean_measurement)
seg_covid_log <- log(seg_covid$cases)

# cbind this to segmented dataframes to plot
seg_aqs_comb <- seg_aqs %>% 
  cbind(seg_aqs_log)

seg_covid_comb <- seg_covid %>% 
  cbind(seg_covid_log)

seg_aqs
seg_covid

saveRDS(seg_aqs,
        file = "./data/seg_aqs.rds")
saveRDS(seg_covid,
        file = "./data/seg_covid.rds")

```

Ok, let's plot to get an idea of variance between log transformations.

```{r bind-one-df}
# convert covid to same date format as aqs, add type
seg_covid_comb_bind <- seg_covid_comb %>% 
  mutate(date = as.Date(date),
         type = "covid") %>% 
  rename(value = cases,
         log = seg_covid_log)

# add type col
seg_aqs_comb_bind <- seg_aqs_comb %>% 
  mutate(type = "aq") %>% 
  rename(value = mean_measurement,
         log = seg_aqs_log)

# rbind df's
seg_bind <- seg_aqs_comb_bind %>% 
  rbind(seg_covid_comb_bind)
```

Plot combined data frame to get legend.

```{r plot-ggplot-one-df}
log_compare_line <- ggplot(seg_bind) +
    geom_line(aes(x = date, y = log, color = type), alpha = 0.8) +
    scale_color_manual(labels = c("Air Quality","COVID Cases"),
      values = c("#0C6291","#A63446")) +
    theme_minimal() +
    labs(x = "", y = "Log scale", 
         title = "COVID Case Count and Air Quality PM2.5 per Day (Log)",
         color = "") +
    theme_minimal() +
    theme(plot.title.position = "plot",
          plot.title = element_text(hjust = 0.5, size = 18),
          axis.text = element_text(size = 12),
          axis.title.y = element_text(size = 14),
          legend.position = c(0.1,0.9))

ggsave(filename = "./images/log_compare_line.png",
       width = 11, height = 7,
       plot = log_compare_line)

log_compare_line
```

